Review Paper Draft

### 2020-06-18 main.tex
Edits to 2.3

In order to clarify and emphasize the importance of this subtle distinction, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section 2 we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \cite{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section 3 we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section 4, we illustrate the concepts described in section 3 via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. These simulation results are used to illustrate the importance of distinguishing between population structure and heterogeneous environmental exposures in penalized regression methods. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.

@@

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. It should be noted that by attempting to reduce the effects of population structure, these methods reduce the effects of environmental confounding by blocking the casual pathway between population structure and observed SNPs.

The genomic control method is a univariate testing procedure which modifies the Cochran-Armitage test for trend by a correction factor. This correction attempts to quantify and remove the effect of population structure on the distribution of the chi-square statistic using markers unrelated to the phenotype of interest and independent of the marker being tested for association \cite{devlin1999genomic, bacanu2000power, wang2009testing}. The genomic control method retains the disadvantages of univariate SNP testing and its power to detect true associations is likely to suffer in the presence of linkage disequilibrium.

The structured association method attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \cite{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall somewhere between these two extremes. By clustering those individuals within in one subpopulation or another we lose information about more high dimensional relatedness.

@@

Like PC-adjustment methods, LMMs rely on a pair-wise similarity matrix to estimate and subsequently correct for relatedness among subjects. However, the term LMM has been used to describe a number of distinct procedures for analyzing structured genetic data which differ in their objectives and methods. There are two main objectives of LMMs used in a structured genetic context: estimating narrow-sense heritability, and estimating individual SNP effects.

@@

Within the category of LMMs that aim to estimate SNP effects are univariate and multivariate approaches. Univariate approaches attempt to estimate univariate SNP effects on phenotype and assess the statistical significance of SNP-phenotype associations while controlling for population structure using a genetic similarity matrix \cite{yu2006unified, kang2010variance, kang2008efficient}. However, as with all univariate testing approaches univariate LMM implementations may still produce spurious associations and biased effect estimates in the presence of LD.

@@

Using simulations, we compare the performance of PC-Lasso and LMM-Lasso in terms of their abilities to accurately estimate SNP effects and in terms of their true sign rates in the presence of varying levels of relatedness and environmental confounding.

Although several studies have shown that LMMs outperform PC-adjustment methods in the univariate framework \cite{wang2013analytical, kang2010variance, zhao2007arabidopsis}, to the best of our knowledge this is the only comprehensive review and comparison of these methods in a penalized multivariate framework and using a simulation scheme motivated by human genetic relatedness and which formalizes the non-genetic confounding mechanism of population stratification.\\

@@

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects belonging to $q \le p$ subpopulations and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$. We denote the standardized version of this genotype matrix as $\bX$. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

In order to clarify and emphasize the importance of this subtle distinction, we present a detailed review of relevant concepts and methods. The remainder of this paper is organized as follows. In section 2 we provide a summary of the causes and consequences of structure in the genetic data of seemingly unrelated individuals and formally define terminology used throughout this paper. We emphasize that confounding due to population structure, frequently cited as a driver of spurious associations \cite{Sillanpaeae2011, sul2018population}, is a function of environmental heterogeneity, where genetic data serves as a proxy for differential environmental exposures. In so doing, we refine the definition of confounding due to population structure in terms of a non-genetic mechanism. In addition, we briefly review historical methods of correcting for population stratification and relatedness. In section 3 we provide a detailed review of the two most prevalent multivariate methods of adjusting for population stratification and relatedness at present: PCA adjusted lasso penalized regression (PC-Lasso) and lasso penalized multivariate LMMs (LMM-Lasso). We review the statistical details of these methods as well as the consequences of the ways in which they attempt to correct for population structure and environmental heterogeneity. In section 4, we illustrate the concepts described in section 3 via simulation studies, where the data-generating mechanism formally differentiates between population structure and environmental heterogeneity. These simulation results are used to illustrate the importance of distinguishing between population structure and heterogeneous environmental exposures in penalized regression methods. In addition, these results identify scenarios where particular methods may be most and least effective in accurately estimating the effect sizes of observed SNPs in the presence of unobserved confounding.
\pb{don't hard-code ``section 3''; use ref and label instead}

@@

Statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. It should be noted that by attempting to reduce the effects of population structure, these methods reduce the effects of environmental confounding by blocking the casual pathway between population structure and observed SNPs \pb{Shouldn't this be: between environment and SNPs?}.

The genomic control method is a univariate testing procedure which modifies the Cochran-Armitage test for trend by a correction factor \pb{genomic control is not specific to trend tests; it can be used for any type of test statistic inflation}. This correction attempts to quantify and remove the effect of population structure on the distribution of the chi-square statistic using markers unrelated to the phenotype of interest and independent of the marker being tested for association \cite{devlin1999genomic, bacanu2000power, wang2009testing}. The genomic control method retains the disadvantages of univariate SNP testing and its power to detect true associations is likely to suffer in the presence of linkage disequilibrium \pb{reference?}.  \pb{This paragraph is a little too negative about genomic control...needs to be softened slightly}

Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \cite{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall somewhere between these two extremes \pb{seems useful to bring in idea of a gradient, or continuously varying relatedness, here}. By clustering those individuals within in one subpopulation or another we lose information about more high dimensional relatedness \pb{I think ``high dimensional relatedness'' is confusing/ambiguous here.  What do you mean by it?}.

@@

Like PC-adjustment methods, LMMs rely on a pair-wise similarity matrix \pb{as we discussed, this is a bit odd to say here} to estimate and subsequently correct for relatedness among subjects. However, the term LMM has been used to describe a number of distinct procedures for analyzing structured genetic data which differ in their objectives and methods. There are two main objectives of LMMs used in a structured genetic context: estimating narrow-sense heritability, and estimating individual SNP effects.

@@

LMMs can also be used to estimate SNP effects, and this can be done in either a univariate or multivariate manner. Univariate approaches attempt to estimate the marginal effect of a SNP on the phenotype and assess the statistical significance of SNP-phenotype associations while controlling for population structure using a genetic similarity matrix \cite{yu2006unified, kang2010variance, kang2008efficient}. However, as with all univariate testing approaches univariate LMM implementations may still produce spurious associations and biased effect estimates in the presence of LD.

@@

In Section~\ref{Sec:res}, we compare the performance of PC-Lasso and LMM-Lasso in terms of their abilities to accurately estimate SNP effects and in terms of their true sign rates in the presence of varying levels of relatedness and environmental confounding. Although several studies have shown that LMMs outperform PC-adjustment methods in the univariate framework \cite{wang2013analytical, kang2010variance, zhao2007arabidopsis}, to the best of our knowledge this is the only comprehensive review and comparison of these methods in a penalized multivariate framework and using a simulation scheme motivated by human genetic relatedness and which formalizes the non-genetic confounding mechanism of population stratification.\\

@@

To describe and compare PC-Lasso and LMM-Lasso, we consider an $n \times p$ matrix of genotype data $\boldsymbol{G}$ for $n$ subjects belonging to $q \le p$ subpopulations and $p$ SNPs, where each $G_{i,j} \in \{ 0, 1, 2 \}$ enumerates the number of allele copies subject $i$ has for SNP $j$ \pb{do we need to explicitly assume q discrete subpopulations here?}. We denote the standardized version of this genotype matrix as $\bX$. We assume that an $n \times 1$ vector of continuous phenotype values, $\by$, can be expressed as the sum of additive SNP effects, environmental confounding, and observational noise and consider the linear mixed model,

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{Sec:res}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%------------------------------%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
